Choosing climate forcing for numerical simulations of the Cordilleran ice sheet

### 2013-03-21 sections/climate.tex
Implement Arjen comments
Trivial comments implemented in the text; others as todo notes.

We force an ice sheet model with temperature and precipitation data from one observational dataset, three global and one regional reanalyses. Here we present this climate data and describe how we use it as a forcing.

@@

WorldClim \citep{data:worldclim} is a high-resolution climate dataset built from observation data. These climate surfaces were generated by spatial interpolation between a large set of weather stations using a 30\,arc\,s-aggregated flavour of hole-filled SRTM data over global land areas. This provides a resolution much higher than attained by general circulation models.

@@

Figure~\ref{fig:temp} shows the spatial distribution of summer (JJA) and winter (DJF) air surface temperatures in WorldClim. Temperatures generally decrease from south to north and regions further inland experience colder winters. It should be noted that the temperature gradient is much stronger in the winter than it is in the summer. At low elevations, temperature get well above zero during the summer months over the entire modelling domain. In other words, there is a strong contrast in seasonality between coastal and inland regions, and regions where mean annual temperatures are well below freezing do experience warm summers.

@@

Figure~\ref{fig:prec} shows the spatial distribution of winter (DJF) and summer (JJA) precipitation rates in WorldClim. Coastal regions generally receive much more precipitation than inland ones as a result of the continuous topographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades. In a glacier mass-balance view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year. Although coastal regions experience most precipitation during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.

In regions like Northern Yukon and Alaska, dry winters and warm summers are unfavourable conditions to accumulation and glacial inception despite of the strongly negative present mean annual temperatures. In order to account for these strong gradients in seasonality, we use monthly means to drive the ice sheet model. Previous runs forced by annual means of temperature and precipitation, not presented here systematically conducted to unrealistic ice build-up in northern Yukon and Alaska under present-day climate while leaving the Coast Mountains free of ice.

The density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous. If good coverage exists in the southern parts of our modelling domain, several hundred kilometres can separate nearby stations in the north \citep{data:worldclim}. A second drawback of the dataset in an ice sheet modelling view is the lack of data on marine surfaces, particularly on the Pacific continental shelf where glaciers have advanced during the last glacial cycle\needref.

@@

Aside WorldClim data, we use surface air temperatures and precipitation rates from three global and one regional climate reanalyses to force the ice sheet model: the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate System Forecast Reanalysis (CFSR) and the North American Regional Reanalysis (NARR). Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado, USA, from their website at \url{http://www.esrl.noaa.gov/psd/} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series. Further information on the data used is gathered in Table \ref{tab:reanalyses}.

The spatial distribution of summer (JJA) air surface temperatures and winter (DJF) precipitation rates from the four reanalyses datasets are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}. Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.

As a mixed product from observations and a circulation model, we believe that reanalysis may perform best in poorly monitored regions such as the northernmost American Cordillera.

@@

Air surface temperature and precipitation rates from the WorldClim data, the NCEP/NCAR, ERA-Interim, CFSR and NARR reanalyses were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the model resolution of 10\,km. In addition, WorldClim data was extrapolated to marine surfaces using a nearest-neighbour approach. These last two steps are not presented on figures~\ref{fig:temp} and~\ref{fig:prec}.

For the CFSR, an alternative forcing was prepared by smoothing the precipitation field in order to correct for artefacts seen in figure~\ref{fig:prec}. This was done by averaging data locally in a circular neighbourhood of 7\,pixels in diameter prior to re-projection. Yet we will show that this smoothing has very limited effect on the ice sheet model outcome. This preprocessing steps were made in GRASS~GIS using scripts available on the first author's website.

>>>>

We force PISM with temperature and precipitation data from one observational dataset, three global and one regional reanalyses. Here we present this climate data and describe how we use it as a forcing.

@@

WorldClim is a high-resolution climate dataset built from observation data \citep{data:worldclim}. These climate surfaces were generated by spatial interpolation between a large set of weather stations over global land areas, using 30\,arc\,s-aggregated hole-filled SRTM elevation data. This provides a resolution much higher than attained by general circulation models.

@@

Figure~\ref{fig:temp} shows the spatial distribution of summer (JJA) and winter (DJF) air surface temperatures in WorldClim. Temperatures generally decrease from south to north and regions further inland experience colder winters. It should be noted that the temperature gradient is much stronger in the winter than it is in the summer. Temperature reach well above zero during the summer months over the entire modelling domain, except of the highest peaks. In other words, there is a strong contrast in seasonality between coastal and inland regions, even and regions where mean annual temperatures are well below freezing do experience warm summers.

@@

\arjen{It is an unusual style to start each sentence with Figure X shows... We should rephrase this later-on.} Figure~\ref{fig:prec} shows the spatial distribution of DJF and JJA precipitation rates in WorldClim. Coastal regions generally receive much more precipitation than inland ones as a result of the continuous topographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades\arjen{orographic precip!}. From a glacier mass-balance point of view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year. Although coastal regions experience most precipitation as snow during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.

In regions like Northern Yukon and Alaska, dry winters and warm summers are unfavourable conditions to ice accumulation and glacial inception despite of the strongly negative present mean annual temperatures. In order to account for these strong gradients in seasonality, we use monthly averages of temperature and precipitation to drive the ice sheet model.

The density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous. If good coverage exists in the southern parts of our modelling domain, several hundred kilometres can separate neighbouring stations in the north \citep{data:worldclim}. A second drawback of the dataset in the scope of this study is the lack of data across oceans, particularly on the Pacific continental shelf where glaciers have advanced during the last glacial cycle\needref.

@@

Apart from the WorldClim data, we use surface air temperatures and precipitation rates from three global and one regional climate reanalyses to force the ice sheet model: the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate System Forecast Reanalysis (CFSR) and the North American Regional Reanalysis (NARR). Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the NOAA/OAR/ESRL Physical Sciences Division, Boulder, Colorado, USA, from their website at \url{http://www.esrl.noaa.gov/psd/} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series. Further information on the data used is gathered in Table \ref{tab:reanalyses}.

The spatial distribution of JJA air surface temperatures and DJF precipitation rates from the four reanalyses datasets are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}. Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.

As a mixed product from observations and circulation modelling, we believe that reanalysis may perform best in sparsely monitored regions such as the northernmost American Cordillera.

@@

Air surface temperature and precipitation rates \arjen{do you really mean rate as in mm/time unit?} from the WorldClim data, the NCEP/NCAR, ERA-Interim, CFSR and NARR reanalyses were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) \arjen{is this general knowledge? I am lost...} and bilinearly interpolated to the model resolution of 10\,km. In addition, WorldClim data was extrapolated to oceans using a nearest-neighbour approach. These last two steps are not presented on figures~\ref{fig:temp} and~\ref{fig:prec}.

For the CFSR, an alternative forcing was prepared by smoothing the precipitation field in order to correct for artefacts seen in figure~\ref{fig:prec}. This was done by averaging data locally in a circular neighbourhood of 7\,pixels in diameter prior to re-projection. Yet we will show that this smoothing has very limited effect on the ice sheet model outcome. This preprocessing steps were made in GRASS~GIS using scripts available on the first author's website\arjen{I am unsure if this is the best way t do it? Were you using principles described by others then you should refer to the litterature instead.}.

### 2013-03-21 sections/intro.tex
Empty sections coming soon...
Added more todonotes in figures
Provided vector location map instead of bitmap
Fixed typos, language and style
Implement Arjen comments
Trivial comments implemented in the text; others as todo notes.

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=8cm]{cordillera-climate-locmap}
	\end{center}
	\caption{Shaded relief map of northern North America. The frame delimits the modelling domain. After ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth}.}
	\label{fig:locmap}
\end{figure}

>>>>

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=8cm]{cordillera-climate-locmap}
	\end{center}
	\caption{Shaded relief map of northern North America. The frame delimits this study's modelling domain. The outlines of the former ice sheets are from \citet{kleman-etal-2010}. The background map was made with ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth}.}
	\label{fig:locmap}
\end{figure}

\emph{Coming soon...}

% Define PISM = Parallel Ice Sheet Model

### 2013-03-21 sections/model.tex
Implement Arjen comments
Trivial comments implemented in the text; others as todo notes.

\begin{table}[t]
	\caption{Values of the positive degree-day model parameters.}
	\label{tab:pdd}
	\vskip4mm
	\centering
	\begin{tabular}{ll}
		\tophline
		Parameter& Value\\
		\middlehline
		degree-day factor for snow & 3\,\unit{mm\,K^{-1}\,day^{-1}}\\
		degree-day factor for ice & 8\,\unit{mm\,K^{-1}\,day^{-1}}\\
		standard deviation of temperatures & 5\,K\\
		refreezing faction & 0.6 \\
		\bottomhline
	\end{tabular}
\end{table}

Monthly temperature and precipitation rate are converted to surface accumulation and ablation by a temperature-index (positive degree-day) model. Accumulation is computed as the amount of precipitation that falls when temperature is below freezing.\ask{Is this true? What is the current threshold in stable 0.5?} Ablation is computed from the number of positive degree-days, defined as the integrand of temperatures above 0\,\degree\,C in one year. Positive degree-days are computed by the \citet{calov-greve-2011} formula to account for day-to-day variability, which is not represented in the monthly climate forcing. The ablation model use different degree-day factors for snow and ice and accounts for refreezing. Complete description of the ablation model is available in PISM documentation \citep[{\url{http://www.pism-docs.org}},][]{web:pism} and model parameters are expressed in table~\ref{tab:pdd}.

\ask[inline]{\emph{Should I include the \citet{calov-greve-2011} integrand formula?}

	The number of positive degree-days is computed annually using a formula derived by \citet{calov-greve-2011} that accounts for day-to-day variability of temperature by assuming a normal distribution of variance $\sigma=5\degree C$ along a central (input) value $T$:

@@

Where $[t_0; t_1]$ is one year and $\mathrm{erfc}$ is the complementary error function.}

A positive degree-day model is a very crude approximation of physical processes occurring at a glacier's surface. Yet we believe that it captures regional patterns of accumulation and ablation that are reasonable enough for our study if accurate climate data is used, as we will show further.\ask{Is this paragraph really needed?}

@@

Ice flow velocities are computed by a heuristic where the Shallow Shelf Approximation (SSA) is used as a ``sliding law'' to a polythermal Shallow Ice Approximation (SIA) \citep{bueler-brown-2009}.

@@

@@

\begin{equation}
	\phi = \left\{\begin{array}{llc}
		10\,\unit{\degree C} & \mathrm{for} &               z<  0\,\unit{m} \\
		z/10 + 10°           & \mathrm{for} &   0\,\unit{m}<z<200\,\unit{m} \\
		30\,\unit{\degree C} & \mathrm{for} & 200\,\unit{m}<z               \\
	\end{array}\right.
\end{equation}

The glacier bed topography responds to the ice load following a regional isostasy model with flexure and relaxation \citep{lingle-clark-1985}\todo{Check \citet{lingle-clark-1985}}.

@@

>>>>

\begin{table}[t]
	\caption{Positive degree-day model parameters.}
	\label{tab:pdd}
	\vskip4mm
	\centering
	\begin{tabular}{ll}
		\tophline
		Parameter& Value\\
		\middlehline
		degree-day factor for snow & 3\,\unit{mm\,K^{-1}\,day^{-1}}\\
		degree-day factor for ice & 8\,\unit{mm\,K^{-1}\,day^{-1}}\\
		standard deviation of temperatures & 5\,K\\
		refreezing faction & 0.6 \\
		\bottomhline
	\end{tabular}
\end{table}

Monthly temperature and precipitation are converted to surface accumulation and ablation by a temperature-index (positive degree-day) model. Accumulation is computed as the amount of precipitation that falls when temperature is below freezing\ask{Is this true? What is the current threshold in stable 0.5?}. Ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\degree\,C in one year. Positive degree-days are calculated by the \citet{calov-greve-2011} formula to accounts for day-to-day temperature variability by assuming a normal distribution of variance $\sigma=5\degree C$ along a central (input) value $T$:

@@

Where $[t_0; t_1]$ is one year and $\mathrm{erfc}$ is the complementary error function. The ablation model employs different degree-day factors for snow and ice and accounts for refreezing (table~\ref{tab:pdd}).

A positive degree-day model is a crude parametrization of physical processes occurring at a glacier's surface. Yet we believe that it captures reasonably accurate patterns of accumulation and ablation for our study area considering the employed climate data.

@@

Ice flow velocities are computed by a heuristic \arjen{not sure what this expresses!!} where the Shallow Shelf Approximation (SSA) is used as a ``sliding law'' to a polythermal Shallow Ice Approximation (SIA) \citep{bueler-brown-2009}.

@@

\arjen[inline]{unless variations in these determine your best fit input dataset, I would not include them in this paper.}

@@

\begin{equation}
	\phi = \left\{\begin{array}{llrll}
		10      & \mathrm{for} &      &z&<  0 \\
		z/10+10 & \mathrm{for} &   0 <&z&<200 \\
		30      & \mathrm{for} & 200 <&z&     \\
	\end{array}\right.
\end{equation}

With $\phi$ in degrees and $z$ in meters above present sea level. The glacier bed topography responds to the ice load following a regional isostasy model with flexure and relaxation \citep{lingle-clark-1985}\todo{Check \citet{lingle-clark-1985}}.

@@

\arjen[inline]{Check Stroeven et al. (2010): "Radiocarbon ages from wood and fossils underlying McConnell till indicate that the McConnell glaciation began after 29,600  300 14C yr BP (Matthews et al., 1990), but glaciers did not advance into major valleys before 26,350  280 14C yr BP (Jackson and Harington, 1991), and a fully developed CIS did not exist until after 23,900  1140 14C yr BP (Lowdon and Blake,1981). Published radiocarbon ages indicate that the St. Elias lobe of the CIS was in retreat by 13,660  180 14C yr BP (Rampton, 1971) and the Selwyn lobe by 12,590  120 14C yr BP (Ward, 1989) (Fig. 2)." and references therein.}

### 2013-03-21 sections/results.tex
Implement Arjen comments
Trivial comments implemented in the text; others as todo notes.

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-cool06}
	\end{center}
	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 6\degC colder than present for each climate forcing.}
	\label{fig:cool06}
\end{figure*}

@@

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-extent}
	\end{center}
	\qiong[inline]{Move ticks on the colorbar. The 1 is confusing.}
	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
	\label{fig:extent}
\end{figure*}

@@

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=9cm]{cordillera-climate-ivolarea}
	\end{center}
	\todo[inline]{Use black-and-white-proof markers and highlight the ``best'' runs.}
	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing.}
	\label{fig:ivolarea}
\end{figure}

>>>>

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-cool06}
	\end{center}
	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 6\degC colder than present for each climate forcing.}
	\label{fig:cool06}
  \arjen[inline]{Why do you inconsistently use K and degrees C?}
\end{figure*}

@@

\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-extent}
	\end{center}
	\qiong[inline]{Move ticks on the colorbar. The 1 is confusing.}
  \arjen[inline]{I think just removing "1" is enough. No other number are in need of moving I think.}
	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
	\label{fig:extent}
\end{figure*}

@@

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=9cm]{cordillera-climate-ivolarea}
	\end{center}
	\todo[inline]{Use black-and-white-proof markers and highlight the ``best'' runs.}
  \arjen[inline]{The shape of these are modulated by the modelling domain. It looks as if the curves are at maximum difference for 5-6 degrees C and are closer again 7-9 degrees C, but perhaps the latter is an artefact from the fact that the ice sheets cannot grow beyond the domain boundaries?}
	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing.}
	\label{fig:ivolarea}
\end{figure}
